# How to Run This Tutorial

Before diving into the analysis, you need two things: the raw sequencing reads and
a working R environment with the required packages. This preliminary page covers both.

## TLDR; Minimal downloads

* A subset of the reads from [zenodo.org/records/20082786](https://zenodo.org/records/20082786)
* Silva138.2 Database for DADA2 from [zenodo.org/records/14169026](https://zenodo.org/records/14169026/files/silva_nr99_v138.2_toGenus_trainset.fa.gz?download=1)
* A script to install the required packages from [GitHub](https://gist.githubusercontent.com/telatin/61fc1dff358a38596add16bcbda247f5/raw/5d8a37307e1cff0b49d8871a0c5515bc5854c9cc/metabar_setup.R)

## The dataset: Getting the raw reads

The dataset used throughout this tutorial is publicly available in the
**European Nucleotide Archive (ENA)** under accession
[PRJEB112055](https://www.ebi.ac.uk/ena/browser/view/PRJEB112055).

### Downloading via the ENA browser

1. Go to <https://www.ebi.ac.uk/ena/browser/view/PRJEB112055>
2. Click the **"Download"** icon and select **"FASTQ files (FTP)"** to get a TSV
   file listing the FTP URLs for every sample.
3. Use `wget` to download the files listed in the TSV:

```{bash}
#| eval: false

# Download the FTP links file from ENA
wget -i <(awk -F'\t' 'NR>1 {split($7,a,";"); for(i in a) print a[i]}' filereport.tsv)
```

---

## Running locally with RStudio

### Install R and RStudio

1. Download and install **R** (≥ 4.3, but latest is preferred) from <https://cran.r-project.org/>
2. Download and install **RStudio Desktop** (free) from
   <https://posit.co/download/rstudio-desktop/>

### Install required packages


Open RStudio and run the following once to install all dependencies:

```{r}
#| eval: false

# Install some tidyverse packages from CRAN
install.packages(c("tidyr","tibble","readr","ggplot2","patchwork"))

# Install the Bioconductor package manager
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# Install DADA2 and phyloseq from Bioconductor
BiocManager::install(c("dada2", "phyloseq", "Biostrings"))
```

Installation can take several minutes the first time. Once complete, verify
everything loaded correctly:

```{r}
#| eval: false

library(dada2)
```

::: {.callout-tip}
## Version check
The tutorial was developed with dada2 ≥ 1.28, phyloseq ≥ 1.44, and
tidyverse ≥ 2.0. If you see unexpected errors, check `packageVersion("dada2")`
and update via `BiocManager::install("dada2")` if needed.
:::

---

## Running on the NBI HPC

::: {.callout-important}
## HPC account required
You need an active NBI HPC account to follow this route. If you do not have one,
contact your line manager or the NBI IT helpdesk to request access.
:::

The NBI HPC provides RStudio through the **Open on Demand** web portal — no
command-line setup required.

### Step-by-step

1. Open your browser and go to <https://ood.hpc.nbi.ac.uk/>
2. Log in with your NBI HPC credentials.
3. In the top navigation bar click **"Interactive Apps"** and select **RStudio**.
4. Choose the compute resources you need (a 8 cores and 64 GB RAM are enough
   for this tutorial) and click **"Launch"**.
5. Wait for the job to start, then click **"Connect to RStudio Server"**.
6. Run the [installation script](https://gist.githubusercontent.com/telatin/61fc1dff358a38596add16bcbda247f5/raw/5d8a37307e1cff0b49d8871a0c5515bc5854c9cc/metabar_setup.R) to install dependencies

### Installing packages on the HPC

The shared R installation on the HPC does **not** include DADA2, phyloseq, or
tidyverse by default. Run the same installation commands as for a local setup —
R will install the packages into your personal library (e.g. `~/r/lib/`):

```{r}
#| eval: false

if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c("dada2", "phyloseq", "Biostrings"))

install.packages("tidyverse")
```
 
