# Setup and Quality Assessment

```{r}
#| label: setup
#| include: false
source("_common.R")
```

## Load libraries

We start by loading the packages needed for the entire analysis. It is good practice
to load all dependencies at the top of the script so that any missing package is caught
immediately.

```{r}
#| label: libraries

library(dada2)      # Core DADA2 functions
library(tidyverse)  # Data manipulation and ggplot2
library(phyloseq)   # Microbiome data structures (used later)
```

## Configure paths

We define all file paths relative to this project's working directory. Using relative
paths makes the analysis portable — moving the project folder will not break the code.

```{r}
#| label: paths

# Directory containing the renamed, trimmed FASTQ files
# (reads are already primer-trimmed and renamed to clean sample names)
path_reads <- "../reads/"

# SILVA reference database for taxonomy assignment (used in Chapter 5)
path_silva <- "../db/silva_nr_v138_train_set.fa.gz"

# Directory for intermediate outputs
path_filtered <- "filtered"
path_precomp  <- "precomputed"

dir.create(path_filtered, showWarnings = FALSE)
dir.create(path_precomp,  showWarnings = FALSE)
```

## Removing primers

DADA2 works on raw reads but after primer removal. 
We used [cutadapt](https://cutadapt.readthedocs.io/en/stable/) to remove the primers.

* Primer forward: `CCTACGGGNGGCWGCAG`
* Primer reverse: `GGACTACHVGGGTATCTAATCC`

The typical command to remove primers using cutadapt is:

```{text}
cutadapt -a FWDPRIMER...RCREVPRIMER -A REVPRIMER...RCFWDPRIMER \
   --discard-untrimmed -m 100:100 -j 8 \
   -o sample1_noprimer_R1.fq.gz -p sample1_noprimer_R2.fq.gz \
   sample1_R1.fq.gz sample1_R2.fq.gz
```

* `-a` and `-A` provide the primers to trim, not the three dots `...` to separate the pairs. The first is applied to the first pair, while the second in the second pair. We provide *both* primers to each end for the case when the amplicon is shorther than the read length (e.g. ITS sequencing)
* `--discard-untrimmed` means not to save reads where the primers were not detected at all
* `-m X:Y` means not to save reads shorter than X (in R1) and Y (in R2), by default we could end up with empty reads
* `-o R1` and `-p R2` will specify the output files
* `-j INT` is for multithreading (in the example 8 cores used)
* the input are the positional arguments at the end (forward and reverse respectively)

## Discover input files

DADA2 processes forward (R1) and reverse (R2) reads separately until the merging step.
We find all FASTQ files matching each pattern and sort them to guarantee that the
forward file at position `i` corresponds to the reverse file at position `i`.

```{r}
#| label: find-files

fnFs <- sort(list.files(path_reads, pattern = "_R1.fastq.gz", full.names = TRUE))
fnRs <- sort(list.files(path_reads, pattern = "_R2.fastq.gz", full.names = TRUE))

cat("Forward read files found:", length(fnFs), "\n")
cat("Reverse read files found:", length(fnRs), "\n")
```

### Extract sample names

Sample names are extracted from the forward file names. Our files follow the naming
convention `<SampleID>_R1.fastq.gz`, so we strip the `_R1.fastq.gz` suffix.

```{r}
#| label: sample-names

sample.names <- sub("_R1\\.fastq\\.gz$", "", basename(fnFs))
head(sample.names, 8)
```

::: {.callout-tip}
## File naming matters
DADA2 assumes the forward and reverse file vectors are in the same order. Using `sort()`
on both ensures correct pairing even when the filesystem returns files in an unexpected
order.
:::

## Quality assessment

Before processing reads it is essential to visualise their quality profiles. This
informs the truncation lengths we will use in the next chapter.

::: {.callout-tip}
## Quick preview online
you can check the quality profile using our [utility website](https://corebio.info/ai-training/examples/fastq.html), without loading R at all.
:::

### Forward reads

```{r}
#| label: qc-forward
#| fig-width: 10
#| fig-height: 6

plotQualityProfile(fnFs[1:6])
```

### Reverse reads

```{r}
#| label: qc-reverse
#| fig-width: 10
#| fig-height: 6

plotQualityProfile(fnRs[1:6])
```

## Interpreting quality profiles

Each panel shows one sample. 
The *green line* is the mean quality score at each
position, the *orange lines* are the quartiles, and the *red line* (at the bottom)
shows the proportion of reads that extend to that position.

What to look for:

- **Drop in mean quality**: reads typically degrade towards the 3′ end. Look for where
  the green line consistently falls below Q30.
- **Reverse reads are usually worse** than forward reads — this is normal for
  paired-end sequencing.
- **Truncation point**: you need to cut reads short enough to remove low-quality bases,
  but long enough so that the forward and reverse reads still overlap when merged.

Note that we also need to evaluate if we will be able to merge the two reads (if this is intended to happen). A typical V3-V4 amplicon is 460bp long, which means that 300bp paired end sequencing leaves a nice overlap to allow merging. If we need to trim a lot of bases so that their sum is going to be less than ~490bp, the overlap might not be sufficient.

### Quality patterns in NextSeq 2000 data


This dataset was sequenced on an **Illumina NextSeq 2000**, which uses a 2-colour
chemistry and reports **binned** (discrete) quality scores rather than the continuous
Phred scores produced by older instruments like MiSeq. You will typically see quality
scores clustering at a small number of levels (e.g. Q2, Q12, Q23, Q37) rather than
spreading smoothly from 0 to 40.

This binning has an important consequence for DADA2's error modelling step — we will
address it in @sec-error-models.

::: {.callout-note}
## Choosing truncation lengths
Based on these profiles, we will truncate:

- **Forward reads** at position **260** (quality remains high to this point)
- **Reverse reads** at position **250** (slightly shorter to remove end degradation)

After truncation, the overlap between the two reads is sufficient for merging the
V3–V4 amplicon (~460 bp expected product; reads of 260 + 250 = 510 bp total gives
~50 bp overlap, which is adequate).
:::
