# Filtering and Trimming

```{r}
#| label: setup
#| include: false
source("_common.R")
```

## Why filter reads?

Raw sequencing reads contain:

- **Low-quality bases** at the 3′ end that increase error rates in denoising
- **Ambiguous bases** (N) that DADA2 cannot handle
- **PhiX spike-in reads** — a control used for calibration during sequencing
- Occasionally, reads contaminated by the **PhiX bacteriophage genome**

The `filterAndTrim()` function removes all of these in a single step and also truncates
reads to a fixed length, which is required for consistent error modelling.

## Define filtered file paths

Here is where we will *save* the filtered reads.

```{r}
#| label: filtered-paths

filtFs <- file.path(path_filtered,
                    paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path_filtered,
                    paste0(sample.names, "_R_filt.fastq.gz"))

# Give each vector element the corresponding sample name for easy indexing later
names(filtFs) <- sample.names
names(filtRs) <- sample.names
```

## Run filterAndTrim

The *filterAndTrim* function allows to filter our reads that should be already free of sequencing primers!

```{r}
#| label: filter-trim
#| cache: true

out <- filterAndTrim(
  fnFs, filtFs,
  fnRs, filtRs,
  truncLen = c(260, 250),   # truncate F reads at 260 bp, R reads at 250 bp
  maxN     = 0,             # discard any read with an ambiguous base
  maxEE    = c(2, 2),       # maximum expected errors per read (F and R separately)
  truncQ   = 10,            # truncate at first base with quality ≤ 10
  rm.phix  = TRUE,          # remove reads matching PhiX genome
  compress = TRUE,          # write gzip-compressed output
  multithread = TRUE        # use all available CPU cores
)
saveRDS(out, file.path(path_precomp, "out.rds"))
```

## Parameter guide

| Parameter | Value | Rationale |
|-----------|-------|-----------|
| `truncLen` | `c(260, 250)` | Keeps high-quality region; must be long enough for reads to overlap after merging |
| `maxN` | `0` | DADA2 requires zero ambiguous bases |
| `maxEE` | `c(2, 2)` | Maximum *expected errors* (sum of error probabilities along the read); 2 is a standard threshold |
| `truncQ` | `10` | Removes the very-low-quality tail before the maxEE calculation |
| `rm.phix` | `TRUE` | Always advisable; PhiX contamination is common in Illumina data |

::: {.callout-tip}
## What is Expected Errors (EE)?
The **expected errors** metric (introduced by Edgar & Flyvbjerg 2015) is more
informative than a simple minimum quality threshold. For a read with quality scores
Q₁, Q₂, …, Qₙ, the EE is:

$$EE = \sum_{i=1}^{n} 10^{-Q_i/10}$$

A read with `maxEE = 2` can have no more than 2 expected sequencing errors in total.
This is more lenient than requiring every base to exceed Q30, while still keeping
read quality high.
:::

## Inspect filtering statistics

```{r}
#| label: filter-stats

# Convert to data frame and add sample names
filter_stats <- as.data.frame(out)
filter_stats$sample   <- rownames(filter_stats)
filter_stats$pct_kept <- round(100 * filter_stats$reads.out / filter_stats$reads.in, 1)

filter_stats[, c("sample", "reads.in", "reads.out", "pct_kept")]
```

```{r}
#| label: filter-summary

cat("Median reads retained:", median(filter_stats$pct_kept), "%\n")
cat("Minimum reads retained:", min(filter_stats$pct_kept), "%\n")
```

A retention rate above **70–80 %** is generally acceptable. If many samples lose more
than 30 % of reads at this step, consider relaxing `maxEE` slightly (e.g. to 5) or
revisiting the primer-trimming step upstream.

## Check filtered files exist

A small sanity check: some samples may have lost all reads (e.g. negative controls
with very low input). DADA2 will error if it tries to open an empty or absent file.

```{r}
#| label: check-filtered

# Keep only samples for which filtered files were actually created
exists_F <- file.exists(filtFs)
exists_R <- file.exists(filtRs)

if (any(!exists_F | !exists_R)) {
  dropped <- sample.names[!exists_F | !exists_R]
  message("The following samples had no reads after filtering and will be excluded: ",
          paste(dropped, collapse = ", "))
  filtFs <- filtFs[exists_F & exists_R]
  filtRs <- filtRs[exists_F & exists_R]
  sample.names <- names(filtFs)
}

cat("Samples remaining:", length(sample.names), "\n")
```
