# Denoising, Merging and Chimera Removal

```{r}
#| label: setup
#| include: false
source("_common.R")

# Load objects produced by earlier chapters
errF <- readRDS(file.path(path_precomp, "errF.rds"))
errR <- readRDS(file.path(path_precomp, "errR.rds"))
out  <- readRDS(file.path(path_precomp, "out.rds"))
```

## Denoising with DADA2

The core DADA2 algorithm uses the error model learned in the previous chapter to
distinguish true biological sequences from sequences that arose through PCR or
sequencing errors. It works by asking: *given the error model, is this slightly
different sequence more likely to be a genuine variant, or an error-derived copy of
a more abundant sequence?*

The result is a set of **Amplicon Sequence Variants (ASVs)**: exact sequences with no
artificial clustering.

```{r}
#| label: denoise
#| cache: true

dadaFs_file <- file.path(path_precomp, "dadaFs.rds")
dadaRs_file <- file.path(path_precomp, "dadaRs.rds")

if (!file.exists(dadaFs_file)) {
  message("Denoising forward reads...")
  dadaFs <- dada(filtFs, err = errF, multithread = TRUE)
  saveRDS(dadaFs, dadaFs_file)
} else {
  dadaFs <- readRDS(dadaFs_file)
}

if (!file.exists(dadaRs_file)) {
  message("Denoising reverse reads...")
  dadaRs <- dada(filtRs, err = errR, multithread = TRUE)
  saveRDS(dadaRs, dadaRs_file)
} else {
  dadaRs <- readRDS(dadaRs_file)
}
```

### What does the dada2 output tell us?

```{r}
#| label: denoise-inspect

# Inspect the first sample's denoising result
dadaFs[[1]]
```

The output reports how many unique sequences were identified in the sample and how many
reads support each one. The ratio of reads to unique sequences gives an indication of
how much error correction was performed.

## Merging paired-end reads

After denoising, forward and reverse reads are merged to reconstruct the full amplicon
sequence. Merging requires a minimum overlap region between the denoised forward and
reverse reads — by default, at least 12 bp.

```{r}
#| label: merge
#| cache: true

mergers_file <- file.path(path_precomp, "mergers.rds")

if (!file.exists(mergers_file)) {
  message("Merging paired reads...")
  mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose = TRUE)
  saveRDS(mergers, mergers_file)
} else {
  mergers <- readRDS(mergers_file)
}
```

::: {.callout-tip}
## Merging failure
If a large proportion of reads fail to merge (visible in the tracking table later),
the most common causes are:

1. **Insufficient overlap** — increase `truncLen` for one or both reads, or reduce it
   if the reads are too short
2. **Primer residues** — verify that primers were fully removed before this analysis
3. **Wrong amplicon size** — the reads must cover the amplicon with some overlap
:::

## Build the sequence table

`makeSequenceTable()` produces a sample × ASV matrix of read counts. This is the
central data object in DADA2.

```{r}
#| label: seqtab

seqtab <- makeSequenceTable(mergers)
cat("Dimensions (samples × ASVs):", dim(seqtab), "\n")
```

### Sequence length distribution

The expected amplicon length for the V3–V4 region is approximately 400–470 bp. Any
sequences outside this range are likely artefacts (non-specific amplification or
chimeras) and will often be removed in subsequent steps.

```{r}
#| label: length-dist
#| fig-width: 8
#| fig-height: 4

seq_lengths <- table(nchar(getSequences(seqtab)))
print(seq_lengths)

# Visualise
barplot(seq_lengths,
        xlab = "Sequence length (bp)",
        ylab = "Number of ASVs",
        main = "ASV length distribution (before chimera removal)",
        las  = 2)
```

## Chimera removal

Chimeric sequences are PCR artefacts created when an incomplete extension product
from one cycle acts as a primer in a subsequent cycle, fusing the 5′ end of one
template with the 3′ end of another. DADA2 identifies chimeras using the
`removeBimeraDenovo()` function, which checks whether each sequence can be
reconstructed from two more-abundant "parent" sequences.

```{r}
#| label: chimera
#| cache: true

seqtab_file <- file.path(path_precomp, "seqtab_nochim.rds")

if (!file.exists(seqtab_file)) {
  message("Removing chimeras...")
  seqtab.nochim <- removeBimeraDenovo(seqtab,
                                       method      = "consensus",
                                       multithread = TRUE,
                                       verbose     = TRUE)
  saveRDS(seqtab.nochim, seqtab_file)
} else {
  seqtab.nochim <- readRDS(seqtab_file)
}

cat("ASVs before chimera removal:", ncol(seqtab), "\n")
cat("ASVs after  chimera removal:", ncol(seqtab.nochim), "\n")
```

```{r}
#| label: chimera-proportion

pct_retained <- round(100 * sum(seqtab.nochim) / sum(seqtab), 1)
cat("Proportion of reads retained after chimera removal:", pct_retained, "%\n")
```

::: {.callout-note}
## Interpreting chimera removal
It is **normal** for a large *number* of unique sequences to be flagged as chimeric —
often 10–30 % of ASVs. However, chimeric reads usually represent a small proportion
of total *reads* (< 10 %), because chimeras are typically rare relative to their
parent sequences. If you lose > 20 % of reads here, double-check that primers were
fully trimmed.
:::

```{r}
#| label: length-dist-nochim
#| fig-width: 8
#| fig-height: 4

seq_lengths_nc <- table(nchar(getSequences(seqtab.nochim)))
print(seq_lengths_nc)

barplot(seq_lengths_nc,
        xlab = "Sequence length (bp)",
        ylab = "Number of ASVs",
        main = "ASV length distribution (after chimera removal)",
        las  = 2)
```

## Tracking reads through the pipeline

A key quality-control step is to track how many reads survive each stage of the
pipeline. Large losses at any single step point to a specific problem.

```{r}
#| label: tracking

getN <- function(x) sum(getUniques(x))

track <- cbind(
  out,
  denoisedF = sapply(dadaFs, getN),
  denoisedR = sapply(dadaRs, getN),
  merged    = sapply(mergers, getN),
  nonchim   = rowSums(seqtab.nochim)
)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR",
                     "merged", "nonchim")
rownames(track) <- sample.names

write.csv(track, "dada2_QC.csv")
track
```

### Visualise read tracking

```{r}
#| label: tracking-plot
#| fig-width: 9
#| fig-height: 5

step_labels <- c(
  input     = "Raw Input",
  filtered  = "Quality Filtered",
  denoisedF = "Denoised (Fwd)",
  denoisedR = "Denoised (Rev)",
  merged    = "Merged Pairs",
  nonchim   = "Non-chimeric"
)

track_long <- as.data.frame(track) %>%
  rownames_to_column("Sample") %>%
  pivot_longer(-Sample, names_to = "Step", values_to = "Reads") %>%
  mutate(Step = factor(Step, levels = names(step_labels)))

ggplot(track_long, aes(x = Step, y = Reads, fill = Step)) +
  geom_boxplot(outlier.size = 1) +
  scale_x_discrete(labels = step_labels) +
  scale_fill_viridis_d() +
  labs(x = "Pipeline Step", y = "Number of Reads",
       title = "Read counts through the DADA2 pipeline") +
  theme_minimal() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

**What to look for:**

- A gradual, smooth decline across steps is expected and healthy
- A sudden large drop at the **filtering** step → truncation lengths may be too aggressive
- A large drop at **merging** → check overlap between reads; check primers are removed
- A large drop at **chimera removal** → investigate whether primers are truly absent
