# Taxonomy Assignment

```{r}
#| label: setup
#| include: false
source("_common.R")
seqtab.nochim <- readRDS(file.path(path_precomp, "seqtab_nochim.rds"))
```

## Overview

Once we have a table of denoised, chimera-free ASV sequences, we can assign each one
a taxonomic identity by comparing it to a reference database. DADA2 uses a
**k-mer–based naive Bayesian classifier** (Wang et al. 2007) implemented in
`assignTaxonomy()`.

The function assigns taxonomy from Kingdom down to Genus, returning a confidence
**bootstrap value** for each level. Assignments with low bootstrap support (< 50 % by
default) are reported as `NA` rather than an unreliable label.

## The SILVA reference database

We use the **SILVA** database (v138), one of the most comprehensive curated 16S/18S
rRNA gene databases. We use the formatted training set provided by the DADA2 team:

| File | Content |
|------|---------|
| `silva_nr_v138_train_set.fa.gz` | Kingdom → Genus classifier |

::: {.callout-note}
## Species-level assignment
Species-level assignment requires a second, separate reference file
(`silva_v138_species_assignment.fa.gz`). It uses exact matching of the full ASV
sequence rather than the Bayesian classifier, and should only be trusted when the
amplicon region is discriminating at the species level. We skip `addSpecies()` in
this tutorial because the file is not included in the dataset, but it can be added
as a straightforward extra step if needed.
:::

## Run taxonomy assignment

```{r}
#| label: assign-taxonomy
#| cache: true

taxa_file <- file.path(path_precomp, "taxa.rds")

if (!file.exists(taxa_file)) {
  message("Assigning taxonomy (this may take several minutes)...")
  taxa <- assignTaxonomy(seqtab.nochim,
                         path_silva,
                         multithread = TRUE,
                         verbose     = TRUE)
  saveRDS(taxa, taxa_file)
} else {
  message("Loading pre-computed taxonomy.")
  taxa <- readRDS(taxa_file)
}
```

## Inspect the taxonomy table

```{r}
#| label: inspect-taxa

# Remove sequence row names for readability
taxa.print <- taxa
rownames(taxa.print) <- NULL
head(taxa.print, 10)
```

The `taxa` object is a matrix with one row per ASV and one column per taxonomic
level. `NA` values indicate the classifier was not sufficiently confident at that
level.

## Classification summary

How many ASVs received a classification at each level?

```{r}
#| label: classification-summary

taxa_df <- as.data.frame(taxa)
levels  <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus")

classif <- data.frame(
  Level      = factor(levels, levels = levels),
  ASVs_classified = sapply(levels, function(lv) sum(!is.na(taxa_df[[lv]]))),
  Total      = nrow(taxa_df)
) %>%
  mutate(Percentage = round(100 * ASVs_classified / Total, 1))

classif
```

```{r}
#| label: classification-plot
#| fig-width: 7
#| fig-height: 4

ggplot(classif, aes(x = Level, y = Percentage, fill = Level)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = paste0(Percentage, "%")),
            vjust = -0.4, size = 3.5) +
  scale_fill_viridis_d() +
  labs(title = "Proportion of ASVs classified at each taxonomic level",
       x     = "Taxonomic level",
       y     = "% of ASVs") +
  theme_minimal() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  ylim(0, 110)
```

::: {.callout-tip}
## Interpreting classification depth
- Near 100 % classification at **Kingdom** and **Phylum** is expected for a clean
  16S dataset
- Classification depth drops progressively at lower levels (Family → Genus) — this
  is normal and reflects the limits of 16S resolution
- Very low Phylum-level classification (< 80 %) may indicate: non-bacterial sequences,
  contamination, or a mismatch between the amplified region and the database
:::
